Demographic Species Distribution Models

### 2019-10-16 manuscript/appendix_a.Rmd
add observation models section appendix

>>>>

# Alternative observation models

Below I provide alternative observation models for count, and presence-only data. occupancy detection observation models can be implemnted using other greta extension packages, but are byond the scope of this manuscript.

## Poisson count data

The relative abundance variable $Lambda_i$ is the parameter of the poisson distribution, so a poisson count model is simply given by:

Notation:
\begin{align}
y_i &\sim Poisson(\Lambda_i)
\end{align}

Code:
```{r poisson_obs}
distribution(y) <- poisson(Lambda)
```
Where `y` is a vector of counts

## Negative binomial count data

Where count data are overdispersed with respect to the expected counts $\Lambda_i$, this overdispersion can be modelled with a negative binomial distribution as:

Notation:
\begin{align}
p_i &= r / (r + \Lambda_i) \\ \nonumber
y_i &\sim Negative Binomial(r, p_i)
\end{align}
Where the negative binomial is parameterised in terms of dispersion parameter $r$ representing the target number of successful trials, and $p_i$ the probability of success in each trial.

Code:
```{r negbin_obs}
p <- r / (r + Lambda)
distribution(y) <- negative_binomial(r, p)
```
Given a positive scalar `r`, and where y is a vector of counts.

## Presence-only data

We assume presence-only data (where records consist of locations where the species was detected) arise from an a thinned, inhomogenous Poisson point process, as described in [@rennerPointProcessModels2015]. Under this definition, the DSDM relative abundance parameter $\Lambda_i$ represents a density of point observations per unit area (assumed homogenous across each site $i$) that *would* be seen if there were no spatial bias in rates of reporting. The expected number of presence-only observations in a location $i$ is therefore given by the product of the relative abundance $\Lambda_i$, the area $a_i$, and a relative measure of the bias in reporting, $\gamma_i$ (also assumed homogenous across $i$). Values of $\gamma_i$ for each location can be provided as data, or estimated during model-fitting via an additional regression submodel against detection covariates. Since $\Lambda_i$ is only identified to within a multiplicative constant ($\alpha$ in Equation 2), the absolute value of $\gamma_i$ is not identified in this model, so any model for $\gamma_i$ should probably not have an intercept term.

Notation:
\begin{align}
y_i &\sim Poisson(a_i\gamma_i\Lambda_i)
\end{align}

Code:
```{r ppm_obs}
distribution(y) <- poisson(a * gamma * Lambda)
```
Where `y` gives the number of presence-only records reported at site $i$

## Integrated models

Specification of integrated models (those with multiple likelihoods for different datatypes) is simple, simply by using multiple `distribution()` statements. For example, a a combined presence-absence and abundance model might be defined as:

```{r integrated_obs}
p <- 1 - exp(-Lambda[pa_locations])
distribution(y_pa) <- bernoulli(p)
distribution(y_count) <- poisson(Lambda[count_locations])
```
Where `pa_locations` and `count_locations` are numeric indices to the sites with to presence-absence and count data respectively, and `y_pa` and `y_count` are the corresponding vectors of observed data.

### 2019-10-16 manuscript/ms.Rmd
add observation models section appendix

>>>>

Since these observation models all relate to $\Lambda_i$, multiple likelihoods could be specified to form an *integrated* DSDM to be fitted to multiple datasets. The integration of multiple liklihoods could enable estimation of parameters such as spatial bias in reporting rates of presence-only data, or to leverage datasets which differ in their spatial and temporal extents or informativeness with respect to growth rates and vital rates.
